Online learning methods for financial time series forecasting

### 2017-12-07 slides/defense-slides.tex
Preparing first changes

\documentclass{beamer}
\usetheme{Warsaw}
%\usecolortheme{lily}
%\usetheme{metropolis}
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{tikz}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

\usebackgroundtemplate{%             declare it
\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
}

@@

\begin{frame}[plain]
\titlepage
\end{frame}
\usebackgroundtemplate{ }

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: economic relations
\item Cointegration and Integration concepts
\item VECM model
\item Proposal
\item Experiments
\item Conclusions
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Motivation}
Most financial time series are non-stationary in nature. This implies that the statistical properties, such as mean and variance of the data change over time. Therefore, financial time series forecasting is still a challenge.
\begin{figure}
\includegraphics[width=0.4\paperwidth]{img/forecast}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}

The hole distribution of the variable does not depend on time, more precisely:
\small
\begin{equation*}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
\end{equation*}
\end{block}
\end{frame}

@@

A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

\begin{frame}
\frametitle{Non-stationary processes}
There are different types of non-stationary time series models often found in economics:
\begin{block}{Deterministic trend}
 Deterministic trend or trend stationary
processes have the following form:
{\color{blue}
\[
y_t = f(t) + \epsilon_{t}
\]}
\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
\end{block}
\end{frame}

\begin{frame}
\frametitle{Non-stationary processes}

\begin{block}{Stochastic trend}
Stochastic trend processes are also called {\color{red}unit root or I(1)} and have the following form:
{\color{blue}
\[
y_t = \mu + y_{t-1} + \epsilon_t
\]}
\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
\end{block}
\end{frame}

\begin{frame}[t,fragile]{Random walk Illustration}
 \begin{itemize}
  \item<1-> The drunkard's walk
  \item<2-> The dog's walk
  %\item<3-> The drunk and dog's walk
 \end{itemize}

\begin{columns}
\begin{column}{.4\linewidth}<1->
\begin{center}
        \begin{tikzpicture}[overlay]
            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
        \end{tikzpicture}
\end{center}
\end{column}
\begin{column}{.08\linewidth}<2->
        \begin{tikzpicture}[overlay]
            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
        \end{tikzpicture}
\end{column}
%\begin{column}{.5\linewidth}<3->
%    \begin{center}
%        \begin{tikzpicture}[overlay]
%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
%        \end{tikzpicture}
%    \end{center}
%\end{column}
\end{columns}
\end{frame}

\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
 \begin{itemize}
  \item The drunk and dog's walk
 \end{itemize}
    \begin{center}
        \begin{tikzpicture}[overlay]
            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
        \end{tikzpicture}
    \end{center}
\end{frame}

\begin{frame}
\hspace*{-4mm}
\includegraphics[width=0.9\paperwidth]{img/motivation.png}
\end{frame}

@@

@@

% Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.
\end{document}

>>>>

\documentclass{beamer}
\usetheme{CambridgeUS}
%\usecolortheme{lily}
%\usetheme{metropolis}
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{tikz}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

%\usebackgroundtemplate{%             declare it
%\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
%   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
%}

@@

\begin{frame}[plain]
\titlepage
\end{frame}
%\usebackgroundtemplate{ }

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: Financial time series and HFT
\item Cointegration and Integration concepts
\item VECM model
\item Proposal
\item Experiments
\item Conclusions
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{HFT: High Frequency Trading}
High frequency trading has motivated computer-driven strategies capable of processing large amount of data in short periods of time.
\begin{figure}
\includegraphics[width=0.7\paperwidth]{img/HFTmarket}
\caption{Percentage of HFT trades of total US Equity Trading. Source: TABB group}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{HFT Revenue}
However, the revenues have fallen dramatically mainly because of more regulation, increasing cost of HFT infrastructure and competition.
\begin{figure}
\includegraphics[width=0.7\paperwidth]{img/HFTmarket2}
\caption{Revenue in the US. Source: TABB group}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{High Frequency data}
High frequency time series are commonly:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Non-stationary
\item Irregularly spaced over time.
%\item Heterogeneity
\item Heteroscedastic
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[width=0.35\paperwidth]{img/forecast}
\end{figure}
\end{column}
\end{columns}
\vspace{5mm}
This imply that classical statistical models can't be used. Therefore, financial time series forecasting is still a challenge.
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}
A strictly stationary time series $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$ is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely:
\small

\begin{equation*}
{\color{blue}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
\end{equation*}
\noindent where $c_1,\dots,c_L$ are constants.
\end{block}
\end{frame}

@@

A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$ 
\end{frame}

\begin{frame}
\frametitle{Stationarity: consequences}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

\begin{frame}
\frametitle{Non-stationary processes}
There are different types of non-stationary time series models often found in economics:
\begin{block}{Stochastic trend}
Stochastic trend processes are also called {\color{red}unit root or I(1)} if they have the following form:
{\color{blue}
\[
y_t = \mu + y_{t-1} + \epsilon_t
\]}
\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
\end{block}
%\begin{block}{Deterministic trend}
% Deterministic trend or trend stationary
%processes have the following form:
%{\color{blue}
%\[
%y_t = f(t) + \epsilon_{t}
%\]}
%\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
%\end{block}
\end{frame}

%\begin{frame}[t,fragile]{Random walk Illustration}
% \begin{itemize}
%  \item<1-> The drunkard's walk
%  \item<2-> The dog's walk
%  %\item<3-> The drunk and dog's walk
% \end{itemize}
% 
%\begin{columns}
%\begin{column}{.4\linewidth}<1->
%\begin{center}
%        \begin{tikzpicture}[overlay]
%            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
%        \end{tikzpicture}
%\end{center}
%\end{column}
%\begin{column}{.08\linewidth}<2->
%        \begin{tikzpicture}[overlay]
%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
%        \end{tikzpicture}
%\end{column}
%%\begin{column}{.5\linewidth}<3->
%%    \begin{center}
%%        \begin{tikzpicture}[overlay]
%%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
%%        \end{tikzpicture}
%%    \end{center}
%%\end{column}
%\end{columns}
%\end{frame}
%
%\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
% \begin{itemize}
%  \item The drunk and dog's walk
% \end{itemize}
%    \begin{center}
%        \begin{tikzpicture}[overlay]
%            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
%        \end{tikzpicture}
%    \end{center}
%\end{frame}

\begin{frame}
\frametitle{Irregularly spaced over time}
\begin{figure}
\includegraphics[width=\paperwidth]{img/hft-ticks-zoom.png}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Heteroscedastic}
%HASTA AQUI
\end{frame}

@@

%\begin{frame}
%\hspace*{-4mm}
%\includegraphics[width=\paperwidth]{img/hft-ticks.png}
%\end{frame}
%

@@

% Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.

\bibliographystyle{amscls/amsalpha}
\bibliography{reference}
\end{document}
